---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

First, we unzip the compressed file, then read the csv file into data frame rawData.

```{r}
unzip("activity.zip")
rawData <- read.csv("activity.csv")
summary(rawData)
str(rawData)
```

Seeing that the date is a factor, let's convert this to a data object.

```{r}
rawData$date <- as.Date(rawData$date, format = "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

To calculate the mean total number of steps taken per day, first lets group by the date, then calculate the total

```{r echo=FALSE, message=FALSE}
library(knitr)

knit_hooks$set(inline = function(x) {
  prettyNum(round(x,1), big.mark=",")
})
```

```{r message=FALSE, warning=FALSE, results='asis', fig.height=4}
library(dplyr)
library(ggplot2)
dateSum <- rawData %>% group_by(date) %>% summarize(totalsteps = sum(steps, na.rm = TRUE))

ggplot(data=dateSum, aes(dateSum$totalsteps)) + 
        geom_histogram(binwidth=2500, 
        fill="darkorchid", 
        col="black",
        alpha = .6) +
        labs(title="Steps Per Day", y="Frequency", x="")
```

```{r}
meanSteps <- mean(dateSum$totalsteps)
medianSteps <- median(dateSum$totalsteps)
meanSteps
medianSteps
```

## What is the average daily activity pattern?

In order to plot the time series of average steps taken in each daily interval, we must regroup the original data by interval, which we previously converted to a factor. 

```{r fig.height=4}
int <- rawData %>% group_by(interval) %>% summarize(avgsteps = mean(steps, na.rm = TRUE))


ggplot(data=int, aes(interval,avgsteps, group=1, color=avgsteps)) + 
        geom_line(size=1) +
        scale_x_continuous(name="Interval",
        breaks=seq(0,2400,200)) +
        labs(title="Average Steps Per Interval", y="Steps") 
```

The maximum average steps are taken during this interval:
```{r}
maxInt <- filter(int,int$avgsteps == max(int$avgsteps) )
maxInt$interval
```

## Imputing missing values

Calculating the number of missing data points in data set:
```{r echo=TRUE}

missing = nrow(rawData) - sum(complete.cases(rawData))
missing
```

Replace NA values with the average number of steps for that particular interval by using ifelse():
```{r echo=TRUE}
fullData <- rawData
fullData$steps <- ifelse(is.na(fullData$steps), int$avgsteps[int$interval %in% fullData$interval], fullData$steps)

summary(fullData)
```

```{r}
fullSum <- fullData %>% group_by(date) %>% summarize(totalsteps = sum(steps, na.rm = TRUE))

ggplot(data=fullSum, aes(fullSum$totalsteps)) + 
        geom_histogram(binwidth=2500, 
        fill="turquoise3", 
        col="black",
        alpha = .6) +
        labs(title="Steps Per Day", y="Frequency", x="")
```

So, look what happens when NAs are replaced by the average interval steps!
```{r}
meanFull <- mean(fullSum$totalsteps)
medianFull <- median(fullSum$totalsteps)
meanFull
medianFull
```
## Are there differences in activity patterns between weekdays and weekends?

Let's add a factor variable called daytype, and give each date a level of weekday or weekend.

```{r}
fullData$daytype <- as.factor(ifelse(weekdays( fullData$date) %in% c("Saturday","Sunday"), "Weekend", "Weekday")) 
head(fullData)
```
Now we'll plot the interval average step data separating weekdays from weekends
```{r}
intPanel <- fullData %>% group_by(interval,daytype) %>% summarize(avgsteps = mean(steps, na.rm = TRUE))

ggplot(data=intPanel, aes(interval,avgsteps, group=1, color=avgsteps)) + 
        geom_line(size=1) +
        scale_x_continuous(name="Interval",
        breaks=seq(0,2400,400)) +
        labs(title="Average Steps Per Interval", y="Steps") +
        facet_grid( daytype ~ .)

```
